The Precambrian paleogeography of Laurentia

### 2021-03-06 Manuscript/Laurentia_Paleogeo_Manuscript.tex
updates prior to final revisions

\subsubsection{Laurentia's initial formation}
Collision between the Superior province and the composite Slave+Rae+Hearne provinces that resulted in the Trans-Hudson orogeny represents a major event in the formation of Laurentia (Fig. \ref{fig:Laurentia_map}; \citealp{Corrigan2009a}). Terminal collision recorded in the Trans-Hudson orogen is estimated to have occured ca. 1.86 to 1.82 Ga based on constraints such as U-Pb dates of monazite grains and zircon rims \citep{Skipton2016a, Weller2017a}. It was preceded by a period of accretionary and collisional orogenesis that is recorded in the constituent provinces and terranes of Laurentia leading up to the terminal collision of the Trans-Hudson orogeny. This overall story of rapid Paleoproterozoic amalgamation of Laurentia's constituent Archean provinces, including the terminal Trans-Hudson orogeny, was synthesized in the seminal \textit{United Plates of America} paper of \citet{Hoffman1988a} and has been refined in the time since --- particularly with additional geochronological constraints. Of most relevance here, are the events that led to the suturing of the major Archean provinces: the Thelon orogen associated with the collision between the Slave and Rae provinces ca. 2.0 to 1.9 Ga \citep{Hoffman1989c}; the Snowbird orogen associated with ca. 1.90 Ga collision between the Rae and Hearne provinces and associated terranes \citep{Berman2007a, Thiessen2020a}; the Nagssugtoqidian orogen due to the ca. 1.86 to 1.84 Ga collision between the Rae and North Atlantic provinces \citep{St-Onge2009a}; and the Torngat orogen resulting from the ca. 1.87 to 1.85 Ga collision of the southern Meta Incognita province (grouped with the Rae province in older compilations) with the North Atlantic province \citep{St-Onge2009a}.

@@

Growth of Laurentia in the Paleoproterozoic also occured through accretionary orogenesis. This accretion occurred within the Wopmay orogen through ca. 1.88 Ga arc-continent collision that led to the accretion of the Hottah terrane (the Calderian orogeny) and the subsequent emplacement of the Great Bear magmatic zone from ca. 1.88 to 1.84 Ga \citep{Hildebrand2009a}. Coeval with the Trans-Hudson orogeny was the peripheral Penokean orogeny during which both microcontinent blocks (the Marshfield terrane) and arc terranes accreted on the southeastern margin of the west Superior province ca. 1.86 to 1.82 Ga (Fig. \ref{fig:Laurentia_map}; \citealp{Schulz2007a}). Firm evidence of the end of the Penokean orogeny comes from the ca. 1.78 Ga undeformed plutons of the East Central Minnesota Batholith \citep{Holm2005a}.

@@

%%%%%%%%%%%%% Acknowledgements %%%%%%%%%%%%%
\footnotesize
\subsection*{Acknowledgements}
This work was supported by NSF CAREER Grant EAR-1847277. The manuscript benefited from reviews from Athena Eyster, David Evans, and Lauri Pesonen as well as discussions and manuscript feedback from Francis Macdonald, Yuem Park, Toby Rivers, Sarah Slotznick and Justin Strauss. Many participants in the Nordic Paleomagnetism Workshop have contributed to the compilation and evaluation of the pole list utilized herein. Particular acknowledgement goes to David Evans for maintaining and distributing the compiled pole lists. GPlates, and in particular the pyGPlates API, was utilized in this work \citep{Muller2018b}. Figures were made using Matplotlib \citep{Hunter2007a} in conjunction with cartopy \citep{Met-Office2010a} and pmagpy \citep{Tauxe2016a} within an interactive Python environment \citep{Perez2007a}. The chapter text as well as code, data, and reconstructions used in this paper are openly available and licensed for any form of reuse with attribution (CC BY 4.0) in this repository: \url{https://github.com/Swanson-Hysell-Group/Laurentia_Paleogeography}.

>>>>

\subsubsection{Laurentia's initial formation}
Collision between the Superior province and the composite Slave+Rae+Hearne provinces that resulted in the Trans-Hudson orogeny represents a major event in the formation of Laurentia (Fig. \ref{fig:Laurentia_map}; \citealp{Corrigan2009a}). Terminal collision recorded in the Trans-Hudson orogen is estimated to have occurred ca. 1.86 to 1.82 Ga based on constraints such as U-Pb dates of monazite grains and zircon rims \citep{Skipton2016a, Weller2017a}. It was preceded by a period of accretionary and collisional orogenesis that is recorded in the constituent provinces and terranes of Laurentia leading up to the terminal collision of the Trans-Hudson orogeny. This overall story of rapid Paleoproterozoic amalgamation of Laurentia's constituent Archean provinces, including the terminal Trans-Hudson orogeny, was synthesized in the seminal \textit{United Plates of America} paper of \citet{Hoffman1988a} and has been refined in the time since --- particularly with additional geochronological constraints. Of most relevance here, are the events that led to the suturing of the major Archean provinces: the Thelon orogen associated with the collision between the Slave and Rae provinces ca. 2.0 to 1.9 Ga \citep{Hoffman1989c}; the Snowbird orogen associated with ca. 1.90 Ga collision between the Rae and Hearne provinces and associated terranes \citep{Berman2007a, Thiessen2020a}; the Nagssugtoqidian orogen due to the ca. 1.86 to 1.84 Ga collision between the Rae and North Atlantic provinces \citep{St-Onge2009a}; and the Torngat orogen resulting from the ca. 1.87 to 1.85 Ga collision of the southern Meta Incognita province (grouped with the Rae province in older compilations) with the North Atlantic province \citep{St-Onge2009a}.

@@

Growth of Laurentia in the Paleoproterozoic also occurred through accretionary orogenesis. This accretion occurred within the Wopmay orogen through ca. 1.88 Ga arc-continent collision that led to the accretion of the Hottah terrane (the Calderian orogeny) and the subsequent emplacement of the Great Bear magmatic zone from ca. 1.88 to 1.84 Ga \citep{Hildebrand2009a}. Coeval with the Trans-Hudson orogeny was the peripheral Penokean orogeny during which both microcontinent blocks (the Marshfield terrane) and arc terranes accreted on the southeastern margin of the west Superior province ca. 1.86 to 1.82 Ga (Fig. \ref{fig:Laurentia_map}; \citealp{Schulz2007a}). Firm evidence of the end of the Penokean orogeny comes from the ca. 1.78 Ga undeformed plutons of the East Central Minnesota Batholith \citep{Holm2005a}.

@@

%%%%%%%%%%%%% Acknowledgements %%%%%%%%%%%%%
\footnotesize
\subsection*{Acknowledgements}
This work was supported by NSF CAREER Grant EAR-1847277. The manuscript benefited from reviews from Athena Eyster, David Evans, and Lauri Pesonen as well as discussions and manuscript feedback from Francis Macdonald, Yuem Park, Toby Rivers, Sarah Slotznick and Justin Strauss. Bruce Buffett provided insights on the implications of a stagnant lid regime on the generation of the geomagnetic field. Many participants in the Nordic Paleomagnetism Workshop have contributed to the compilation and evaluation of the pole list utilized herein. Particular acknowledgement goes to David Evans for maintaining and distributing the compiled pole lists. GPlates, and in particular the pyGPlates API, was utilized in this work \citep{Muller2018b}. Figures were made using Matplotlib \citep{Hunter2007a} in conjunction with cartopy \citep{Met-Office2010a} and pmagpy \citep{Tauxe2016a} within an interactive Python environment \citep{Perez2007a}. The chapter text as well as code, data, and reconstructions used in this paper are openly available and licensed for any form of reuse with attribution (CC BY 4.0) in this repository: \url{https://github.com/Swanson-Hysell-Group/Laurentia_Paleogeography}.
